Windowed least-squares model reduction for dynamical systems

### 2019-10-09 main.tex
updated appendix

\section{Selection of Sampling Points}
To construct the sampling point matrix used for hyper-reduction in the second numerical experiment, we employ q-sampling~\cite{qdeim_drmac} and the sample 
mesh concept~\cite{carlberg_gnat}. Algorithm~\ref{alg:qdeim} outlines the steps used in the second numerical experiment to compute the sampling points.

@@

\section{Derivation of the Euler--Lagrange Equations}\label{appendix:eulerlagrange}
This section details the derivation of the Euler Lagrange equations. To this end, we consider the generic functional of the form,

@@

\mathcal{J}: (\statey,\stateyDot) \mapsto  \int_a^b \integrand(\statey(t),\stateyDot(t),t)  dt,

@@

where $\integrand: \RR{M} \times \RR{M} \times[a,b] \rightarrow  \RRplus$ (for arbitrary $M$) with $\integrand: (\stateyDiscrete,\stateyDotDiscrete,\timeDummy) \mapsto \integrand(\stateyDiscrete,\stateyDotDiscrete,\timeDummy)$. We now introduce the function $\statez: [a,b] \rightarrow \RR{M}$ that is implicitly defined as an extremum of~\eqref{eq:el_gen} subject to the boundary condition $\statez(a) = \statez_a$. 
We introduce an arbitrary function $\variation : [a,b] \rightarrow \RR{M}$ with the boundary condition $\variationArg{a} = \bz$ and define a variation from the extremal by,

@@

We note that $\statezBar$ satisfies the same boundary condition $\statez$ since $\variationArg{a} = \bz$.

@@

\int_a^b \bigg[ \frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t\big) \frac{\partial \statezBar }{\partial \delta}(t,\epsilon)  + \frac{\partial \integrand}{\partial \stateyDotDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \frac{\partial \statezDotBar}{\partial \delta} (t,\epsilon)\bigg]dt . $$

@@

= \int_a^b \bigg[ \frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon),t \big) \variation(t)  + \frac{\partial \integrand}{\partial \stateyDotDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \variationDot(t) \bigg]dt. $$

@@

=  \int_a^b  \bigg[\frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \variation(t) - \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \bigg) \variation(t) \bigg]dt + \\ \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statezBar(b,\epsilon),\statezDotBar(b,\epsilon) ,b \big) \variation(b)  - \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statezBar(a,\epsilon),\statezDotBar(a,\epsilon) ,a\big) \variation(a) .

@@

Noting that $\variationArg{a} = \bz$,

@@

=  \int_a^b  \bigg[\frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) , t \big) \variation(t) - \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \bigg) \variation(t) \bigg]dt + \\ \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statezBar(b,\epsilon),\statezDotBar(b,\epsilon) ,b \big) \variation(b) . 

@@

  \int_a^b  \bigg[\frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon),t \big) \variation(t) - \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon),t \big) \bigg) \variation(t) \bigg]dt + \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statezBar(b,\epsilon),\statezDotBar(b;\epsilon),b \big) \variation(b)  = 0. 

@@

  \int_a^b  \bigg[\frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statez(t),\statezDot(t),t \big) \variation(t) - \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statez(t),\statezDot(t),t \big) \bigg) \variation(t) \bigg]dt + \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statez(b),\statezDot(b) ,b\big) \variation(b)  = 0. 

@@

\bigg[ \frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statez(t),\statezDot(t) ,t\big) \bigg]^T - \bigg[ \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statez(t),\statezDot(t) \big) \bigg) \bigg]^T = \bz  

@@

\section{Evaluation of gradients in the Euler--Lagrange Equations for \methodAcronym\ with \spatialAcronym\ trial subspaces}\label{appendix:vector_calc}
We now derive the specific form of the Euler--Lagrange equations for the \methodAcronym\ formulation with \spatialAcronym\ trial subspaces. Without loss of generality, we present the derivation for a single window $t \in [0,T]$ with a constant basis $\basisspace$ and weighting matrix $\stweightingMat$.  
To obtain the specific form of the Euler--Lagrange equations for the \methodAcronym\ formulation, we need to evaluate the gradients in~\eqref{eq:el1} for the integrand,
\begin{align*}\label{eq:integrand_apx}
 \minintegrand & \vcentcolon
(\genstateyDiscreteArgnt{}, \genstateyDiscreteDotArgnt{} ,\timeDummy) \mapsto \frac{1}{2} \big[
\basisspace \genstateyDiscreteDotArgnt{} - \velocity(\basisspace \genstateyDiscreteArgnt{}
+ \stateIntercept,\timeDummy) \big]^T \stweightingMatArg{} \big[
\basisspace \genstateyDiscreteDotArgnt{}  - \velocity(\basisspace \genstateyDiscreteArgnt{} +
\stateIntercept,t) \big], \\ & 
\vcentcolon \RR{\romdim} \times \RR{\romdim} \times [0,T]
 \rightarrow \RR{} .  
\end{align*}
To evaluate the gradients, we first expand $\minintegrand$:
\begin{align*}
 \minintegrand(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy)  &= \frac{1}{2} \big[ \basisspace \genstateyDiscreteDotArgnt{}  - \velocity(\basisspace \genstateyDiscreteArgnt{} + \stateIntercept,\timeDummy) \big]^T \stweightingMatArg{} \big[ \basisspace \genstateyDiscreteDotArgnt{} - \velocity(\veloargsromy) \big] \\ 
 &= \frac{1}{2}\big[\basisspace \genstateyDiscreteDotArgnt{}  \big]^T  \stweightingMatArg{}  \big[\basisspace \genstateyDiscreteDotArgnt{} \big]  - \frac{1}{2}\big[ \velocity(\veloargsromy) \big]^T \stweightingMatArg{}  \big[\basisspace \genstateyDiscreteDotArgnt{}\big] - \frac{1}{2} \big[ \basisspace \genstateyDiscreteDotArgnt{} \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big]  \\ &+ \frac{1}{2}\big[\velocity(\veloargsromy) \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big].
\end{align*}
Since $\stweightingMat$ is symmetric,
\begin{equation}\label{eq:int_expand}
 \minintegrand(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy)  = \frac{1}{2}\big[\basisspace \genstateyDiscreteDotArgnt{}  \big]^T  \stweightingMatArg{}  \big[ \basisspace \genstateyDiscreteDotArgnt{} \big]  -  \big[ \basisspace \genstateyDiscreteDotArgnt{} \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big]  + \frac{1}{2}\big[\velocity(\veloargsromy) \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big].
\end{equation}
For notational purposes, we write the above as,
$$  \minintegrand(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) =  \minintegrand_1(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) +  \minintegrand_2(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) +  \minintegrand_3(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy),$$
where
\begin{align*}
&\minintegrand_1(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) \equiv  \frac{1}{2}\big[\basisspace \genstateyDiscreteDotArgnt{}  \big]^T  \stweightingMatArg{}  \big[ \basisspace \genstateyDiscreteDotArgnt{} \big], \\
&   \minintegrand_2(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) \equiv   -  \big[ \basisspace \genstateyDiscreteDotArgnt{} \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big], \\
&  \minintegrand_3(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) \equiv  \frac{1}{2}\big[\velocity(\veloargsromy) \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big] . 
\end{align*}
Constructing the Euler--Lagrange equations for this functional $\minintegrand$ requires evaluating the derivatives $\frac{\partial \minintegrand}{\partial \genstateyDiscrete}$ and $\frac{\partial \minintegrand}{\partial \genstateyDiscreteDot}$. We start by evaluating $\frac{\partial \minintegrand}{\partial \genstateyDiscrete}$ and go term by term.

@@

&= -[\genstateyDiscreteDotArgnt{}]^T \basisspace^T \stweightingMatArg{} \frac{\partial \velocity}{\partial \stateyDiscrete}  \basisspace.

@@

@@

\EP{Here}
Combining all terms, the Euler--Lagrange equation reads,

@@

 \frac{\partial \minintegrand}{\partial \genstateyDiscrete}(\genstateArg{n}{t},\genstateDotArg{n}{t},t)  - \frac{d}{dt} \bigg[ \frac{\partial \minintegrand}{\partial \genstateyDiscreteDotArgnt{}} (\genstateArg{n}{t},\genstateDotArg{n}{t},t) \bigg] =  - \big[ \basisspace \genstateDotArg{n}{t}  \big]^T \stweightingMatArg{n} \big[  \frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsromn) \big] \basisspace \\ + [\velocity(\veloargsromn) ]^T \stweightingMatArg{n} \big[ \frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsromn) \big] \basisspace - \frac{d}{dt} \bigg[  [\genstateDotArg{n}{t} ]^T  \basisspace^T \stweightingMatArg{n} \basisspace - [\velocity(\veloargsromn) ]^T \stweightingMatArg{n} \basisspace \bigg]  = \boldsymbol 0.

@@

$$ - \bigg( \big[ \basisspace \genstateDotArg{n}{t} \big]^T  -  [\velocity(\veloargsromn)] ^T \bigg) \stweightingMatArg{n}
 \frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn) \basisspace - \frac{d}{dt} \bigg[  [\genstateDotArg{n}{t}]^T \basisspace^T \stweightingMatArg{n} \basisspace -  
[\velocity(\veloargsromn) ]^T \stweightingMatArg{n} \basisspace  \bigg] = \boldsymbol 0.$$

@@

$$ -  \basisspace^T \big[ \frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn)  \big]^T \stweightingMatArg{n} \bigg( \big[ \basisspace \genstateDotArg{n}{t}\big]  -  \velocity(\veloargsromn)  \bigg) -  \frac{d}{dt} \bigg[  \basisspace^T \stweightingMatArg{n} \basisspace \genstateDotArg{n}{t}  - \basisspace^T \stweightingMatArg{n} \velocity(\veloargsromn)   \bigg] = \bz.

@@

$$ -  \basisspace^T \big[ \frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn)  \big]^T \stweightingMatArg{n} \bigg(  \basisspace \genstateDotArg{n}{t}  -  \velocity(\veloargsromn) \bigg) -  \basisspace^T \stweightingMatArg{n} \frac{d}{dt} \bigg[   \basisspace \genstateDotArg{n}{t}  - \velocity(\veloargsromn) \bigg] = \bz. $$
Gathering terms and multiplying by negative one, the final form of the Euler Lagrange equations are obtained,
$$ \bigg[\basisspace^T \big[\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn) \big]^T \stweightingMatArg{n} + \basisspace^T \stweightingMatArg{n} \frac{d}{dt} \bigg] \bigg(  \basisspace \genstateDotArg{n}{t} -  \velocity(\veloargsromn) \bigg) = \bz. $$
This is the \methodAcronym-ROM. Note that this is a second order equation and can be written as two separate first order equations. Equation~\ref{eq:clspg_2ord} can be written equivalently as two first order equations by introducing an auxiliary ``costate" variable. Defining the costate as,

@@

\adjoint^n(t) \defeq  \genstateDotArg{n}{t}  -  \mass^{-1} \basisspace^T \stweightingMatArg{n} \velocity(\veloargsromn) ,

@@

we can manipulate Eq.~\ref{eq:clspg_2ord} as follows: %First, to enable hyper-reduction later on, we leverage the fact that $\stweightingMat^2 = \stweightingMat$,

@@

 \bigg[\basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn) \bigg]^T \stweightingMatArg{n} \bigg( \mathbf{I} - \basisspace [\massArgnt{n}]^{-1} \basisspace^T\stweightingMatArg{n} + \basisspace [\massArgnt{n}]^{-1} \basisspace^T \stweightingMatArg{n} \bigg)  + \basisspace^T \stweightingMatArg{n}  \frac{d}{dt} \bigg] \bigg(  \basisspace \dot{\genstate}^n   -  \velocity(\veloargsromn) \bigg) = \bz.
Pulling out the term multiplied by the positive portion of $\basisspace [\massArgnt{n}]^{-1} \basisspace^T \stweightingMatArg{n}$,

@@

\bigg[\basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete} ( \veloargsromn) \bigg]^T \stweightingMatArg{n}\bigg( \mathbf{I} - \basisspace [\massArgnt{n}]^{-1}  \basisspace^T  \stweightingMatArg{n} \bigg)  + \\ \basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsromn) \bigg]^T \stweightingMatArg{n} \basisspace [\massArgnt{n}]^{-1}  \basisspace^T \stweightingMatArg{n} +   \basisspace^T \stweightingMatArg{n} \frac{d}{dt} \bigg] \bigg(  \basisspace \genstateDotArg{n}{t}   -  \velocity(\veloargsromn) \bigg) = \bz.

@@

\basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn) \bigg]^T \stweightingMatArg{n}\bigg( \mathbf{I} - \basisspace [\massArgnt{n}]^{-1}  \basisspace^T \stweightingMatArg{n} \bigg)  \bigg(  \basisspace \genstateDotArg{n}{t}   -  \velocity(\veloargsromn) \bigg)  + \\  
\bigg[ \basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn) \bigg]^T \stweightingMatArg{n} \basisspace [\massArgnt{n}]^{-1} \basisspace^T \stweightingMatArg{n} +   \basisspace^T \stweightingMatArg{n} \frac{d}{dt} \bigg] \bigg(  \basisspace \genstateDotArg{n}{t}  -  \velocity(\veloargsromn) \bigg) = \bz.

@@

Pulling $\mass^{-1} \basisspace^T \stweightingMatArg{n}$ inside the parenthesis on the second term,

@@

\basisspace^T \bigg[\frac{\partial \velocity}{\partial \state^n}(\veloargsromn) \bigg]^T \stweightingMatArg{n}\bigg( \mathbf{I} - \basisspace[\massArgnt{n}]^{-1} \basisspace^T \stweightingMatArg{n} \bigg)  \bigg(  \basisspace \genstateDotArg{n}{t}   -  \velocity(\veloargsromn) \bigg)  + \\  

>>>>

\section{Selection of sampling points}
To construct the sampling point matrix used for hyper-reduction in the second numerical experiment, we employ q-sampling~\cite{qdeim_drmac} and the sample 
mesh concept~\cite{carlberg_gnat}. Algorithm~\ref{alg:qdeim} outlines the steps used in the second numerical experiment to compute the sampling points.

@@

\section{Derivation of the Euler--Lagrange equations}\label{appendix:eulerlagrange}
This section details the derivation of the Euler--Lagrange equations. To this end, we consider the generic functional of the form,

@@

\mathcal{J}: (\statey,\stateyDot) \mapsto  \int_0^T \integrand(\statey(t),\stateyDot(t),t)  dt,

@@

where $\integrand: \RR{M} \otimes \timeSpaceArg{} \times \RR{M} \otimes \timeSpaceArg{} \times[0,T] \rightarrow  \RRplus$ (for arbitrary $M$) with $\integrand: (\stateyDiscrete,\stateyDotDiscrete,\timeDummy) \mapsto \integrand(\stateyDiscrete,\stateyDotDiscrete,\timeDummy)$. We now introduce the function $\statez: [0,T] \rightarrow \RR{M}$ that is implicitly defined as an extremum of~\eqref{eq:el_gen} subject to the boundary condition $\statez(0) = \statez_0$. 
We introduce an arbitrary function $\variation : [0,T] \rightarrow \RR{M}$ with the boundary condition $\variationArg{0} = \bz$ and define a variation from the extremal by,

@@

We note that $\statezBar$ satisfies the same boundary condition $\statez$ since $\variationArg{0} = \bz$.

@@

\int_0^T \bigg[ \frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t\big) \frac{\partial \statezBar }{\partial \delta}(t,\epsilon)  + \frac{\partial \integrand}{\partial \stateyDotDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \frac{\partial \statezDotBar}{\partial \delta} (t,\epsilon)\bigg]dt . $$

@@

= \int_0^T \bigg[ \frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon),t \big) \variation(t)  + \frac{\partial \integrand}{\partial \stateyDotDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \variationDot(t) \bigg]dt. $$

@@

=  \int_0^T  \bigg[\frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \variation(t) - \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \bigg) \variation(t) \bigg]dt + \\ \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statezBar(T,\epsilon),\statezDotBar(T,\epsilon) ,T \big) \variation(T)  - \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statezBar(0,\epsilon),\statezDotBar(0,\epsilon) ,0\big) \variation(0) .

@@

Noting that $\variationArg{0} = \bz$,

@@

=  \int_0^T  \bigg[\frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) , t \big) \variation(t) - \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon) ,t \big) \bigg) \variation(t) \bigg]dt + \\ \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statezBar(T,\epsilon),\statezDotBar(T,\epsilon) ,T \big) \variation(T) . 

@@

  \int_0^T  \bigg[\frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon),t \big) \variation(t) - \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statezBar(t,\epsilon),\statezDotBar(t,\epsilon),t \big) \bigg) \variation(t) \bigg]dt + \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statezBar(T,\epsilon),\statezDotBar(T,\epsilon),T \big) \variation(T)  = 0. 

@@

  \int_0^T  \bigg[\frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statez(t),\statezDot(t),t \big) \variation(t) - \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statez(t),\statezDot(t),t \big) \bigg) \variation(t) \bigg]dt + \frac{\partial \integrand}{\partial \stateyDotDiscrete} \big(\statez(T),\statezDot(T) ,T\big) \variation(T)  = 0. 

@@

\bigg[ \frac{\partial \integrand  }{\partial \stateyDiscrete } \big(\statez(t),\statezDot(t) ,t\big) \bigg]^T - \bigg[ \frac{d}{dt}\bigg( \frac{\partial \integrand}{\partial \stateyDotDiscrete }  \big(\statez(t),\statezDot(t),t \big) \bigg) \bigg]^T = \bz  

@@

\section{Evaluation of gradients in the Euler--Lagrange equations for \methodAcronym\ with \spatialAcronym\ trial subspaces}\label{appendix:vector_calc}
We now derive the specific form of the Euler--Lagrange equations for the \methodAcronym\ formulation with \spatialAcronym\ trial subspaces. Without loss of generality, we present the derivation for a single window $t \in [0,T]$ with a constant basis $\basisspace$ and weighting matrix $\stweightingMat$.  
To obtain the specific form of the Euler--Lagrange equations for the \methodAcronym\ formulation, we need to evaluate the gradients in~\eqref{eq:euler_lagrange_appendix} for the integrand,
\begin{align*}\label{eq:integrand_apx}
 \minintegrand & \vcentcolon
(\genstateyDiscreteArgnt{}, \genstateyDiscreteDotArgnt{} ,\timeDummy) \mapsto \frac{1}{2} \big[
\basisspace \genstateyDiscreteDotArgnt{} - \velocity(\basisspace \genstateyDiscreteArgnt{}
+ \stateIntercept,\timeDummy) \big]^T \stweightingMatArg{} \big[
\basisspace \genstateyDiscreteDotArgnt{}  - \velocity(\basisspace \genstateyDiscreteArgnt{} +
\stateIntercept,t) \big], \\ & 
\vcentcolon \RR{\romdim} \times \RR{\romdim} \times [0,T]
 \rightarrow \RR{} .  
\end{align*}
To evaluate the gradients, we first expand $\minintegrand$:
\begin{align*}
 \minintegrand(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy)  &= \frac{1}{2} \big[ \basisspace \genstateyDiscreteDotArgnt{}  - \velocity(\basisspace \genstateyDiscreteArgnt{} + \stateIntercept,\timeDummy) \big]^T \stweightingMatArg{} \big[ \basisspace \genstateyDiscreteDotArgnt{} - \velocity(\veloargsromy) \big] \\ 
 &= \frac{1}{2}\big[\basisspace \genstateyDiscreteDotArgnt{}  \big]^T  \stweightingMatArg{}  \big[\basisspace \genstateyDiscreteDotArgnt{} \big]  - \frac{1}{2}\big[ \velocity(\veloargsromy) \big]^T \stweightingMatArg{}  \big[\basisspace \genstateyDiscreteDotArgnt{}\big] - \frac{1}{2} \big[ \basisspace \genstateyDiscreteDotArgnt{} \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big]  \\ &+ \frac{1}{2}\big[\velocity(\veloargsromy) \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big].
\end{align*}
Since $\stweightingMat$ is symmetric,
\begin{equation}\label{eq:int_expand}
 \minintegrand(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy)  = \frac{1}{2}\big[\basisspace \genstateyDiscreteDotArgnt{}  \big]^T  \stweightingMatArg{}  \big[ \basisspace \genstateyDiscreteDotArgnt{} \big]  -  \big[ \basisspace \genstateyDiscreteDotArgnt{} \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big]  + \frac{1}{2}\big[\velocity(\veloargsromy) \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big].
\end{equation}
For notational purposes, we write the above as,
$$  \minintegrand(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) =  \minintegrand_1(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) +  \minintegrand_2(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) +  \minintegrand_3(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy),$$
where
\begin{align*}
&\minintegrand_1(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) \equiv  \frac{1}{2}\big[\basisspace \genstateyDiscreteDotArgnt{}  \big]^T  \stweightingMatArg{}  \big[ \basisspace \genstateyDiscreteDotArgnt{} \big], \\
&   \minintegrand_2(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) \equiv   -  \big[ \basisspace \genstateyDiscreteDotArgnt{} \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big], \\
&  \minintegrand_3(\genstateyDiscrete,\genstateyDiscreteDot,\timeDummy) \equiv  \frac{1}{2}\big[\velocity(\veloargsromy) \big]^T \stweightingMatArg{} \big[ \velocity(\veloargsromy) \big] . 
\end{align*}
Constructing the Euler--Lagrange equations for this functional $\minintegrand$ requires evaluating the derivatives $\frac{\partial \minintegrand}{\partial \genstateyDiscrete}$ and $\frac{\partial \minintegrand}{\partial \genstateyDiscreteDot}$. We start by evaluating $\frac{\partial \minintegrand}{\partial \genstateyDiscrete}$ and go term by term.

@@

&= -[\genstateyDiscreteDotArgnt{}]^T \basisspace^T \stweightingMatArg{} \frac{\partial \velocity}{\partial \stateyDiscrete}  \basisspace,

@@

where we have suppressed the arguments of the Jacobian for simplicity; e.g., formally 
$$\frac{\partial \velocity}{\partial \stateyDiscrete} : (\statewDiscrete,\timeDummy) \mapsto \frac{ \partial \velocity }{ \partial \stateyDiscrete} (\statewDiscrete, \timeDummy).$$ 

@@

Combining all terms and evaluating at $(\genstateArg{}{t},\genstateDotArg{}{t},t)$,

@@

 \frac{\partial \minintegrand}{\partial \genstateyDiscrete}(\genstateArg{}{t},\genstateDotArg{}{t},t)  - \frac{d}{dt} \bigg[ \frac{\partial \minintegrand}{\partial \genstateyDiscreteDotArgnt{}} (\genstateArg{}{t},\genstateDotArg{}{t},t) \bigg] =  - \big[ \basisspace \genstateDotArg{}{t}  \big]^T \stweightingMatArg{} \big[  \frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsrom) \big] \basisspace + \\  [\velocity(\veloargsrom) ]^T \stweightingMatArg{} \big[ \frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsrom) \big] \basisspace - \frac{d}{dt} \bigg[  [\genstateDotArg{}{t} ]^T  \basisspace^T \stweightingMatArg{} \basisspace - [\velocity(\veloargsrom) ]^T \stweightingMatArg{} \basisspace \bigg]  = \boldsymbol 0.

@@

- \bigg( \big[ \basisspace \genstateDotArg{}{t} \big]^T  -  [\velocity(\veloargsrom)] ^T \bigg) \stweightingMatArg{}
 \frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsrom) \basisspace -  \\
\frac{d}{dt} \bigg[  [\genstateDotArg{}{t}]^T \basisspace^T \stweightingMatArg{} \basisspace -  
[\velocity(\veloargsrom) ]^T \stweightingMatArg{} \basisspace  \bigg] = \boldsymbol 0.

@@

$$ -  \basisspace^T \big[ \frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsrom)  \big]^T \stweightingMatArg{} \bigg( \big[ \basisspace \genstateDotArg{}{t}\big]  -  \velocity(\veloargsrom)  \bigg) -  \frac{d}{dt} \bigg[  \basisspace^T \stweightingMatArg{} \basisspace \genstateDotArg{}{t}  - \basisspace^T \stweightingMatArg{} \velocity(\veloargsrom)   \bigg] = \bz.

@@

$$ -  \basisspace^T \big[ \frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsrom)  \big]^T \stweightingMatArg{ } \bigg(  \basisspace \genstateDotArg{ }{t}  -  \velocity(\veloargsrom) \bigg) -  \basisspace^T \stweightingMatArg{ } \frac{d}{dt} \bigg[   \basisspace \genstateDotArg{}{t}  - \velocity(\veloargsrom) \bigg] = \bz. $$
Gathering terms and multiplying by negative one, the final form of the Euler--Lagrange equations are obtained,
\begin{equation}\label{eq:el_secondorder}
 \bigg[\basisspace^T \big[\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsrom) \big]^T \stweightingMatArg{} + \basisspace^T \stweightingMatArg{} \frac{d}{dt} \bigg] \bigg(  \basisspace \genstateDotArg{}{t} -  \velocity(\veloargsrom) \bigg) = \bz.
This is the \methodAcronym-ROM. Note that this is a second order equation and can be written as two separate first order equations. Defining the ``costate" as,

@@

\adjointArg{}{\timeDummy}  : \timeDummy \mapsto \genstateDotArg{}{\timeDummy}  -  \mass^{-1} \basisspace^T \stweightingMatArg{} \velocity(\basisspace \genstateArg{}{t} + \stateIntercept, \timeDummy) ,

@@

we can manipulate Eq.~\eqref{eq:el_secondorder} as follows: %First, to enable hyper-reduction later on, we leverage the fact that $\stweightingMat^2 = \stweightingMat$,

@@

@@

\bigg[\basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsrom) \bigg]^T \stweightingMatArg{} \bigg( \mathbf{I} - \basisspace [\massArgnt{}]^{-1} \basisspace^T\stweightingMatArg{} + \basisspace [\massArgnt{}]^{-1} \basisspace^T \stweightingMatArg{} \bigg)  + \basisspace^T \stweightingMatArg{}  \frac{d}{dt} \bigg] \\ 
\bigg(  \basisspace \genstateDotArg{}{t}   -  \velocity(\veloargsrom) \bigg) = \bz.
Pulling out the term multiplied by the positive portion of $\basisspace [\massArgnt{}]^{-1} \basisspace^T \stweightingMatArg{}$,
\bigg[\basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete} ( \veloargsrom) \bigg]^T \stweightingMatArg{}\bigg( \mathbf{I} - \basisspace [\massArgnt{}]^{-1}  \basisspace^T  \stweightingMatArg{} \bigg)  + \\ \basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsrom) \bigg]^T \stweightingMatArg{} \basisspace [\massArgnt{}]^{-1}  \basisspace^T \stweightingMatArg{} +   \basisspace^T \stweightingMatArg{} \frac{d}{dt} \bigg] \bigg(  \basisspace \genstateDotArg{}{t}   -  \velocity(\veloargsrom) \bigg) = \bz.

@@

\basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsrom) \bigg]^T \stweightingMatArg{}\bigg( \mathbf{I} - \basisspace [\massArgnt{}]^{-1}  \basisspace^T \stweightingMatArg{} \bigg)  \bigg(  \basisspace \genstateDotArg{}{t}   -  \velocity(\veloargsrom) \bigg)  + \\  
\bigg[ \basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsrom) \bigg]^T \stweightingMatArg{} \basisspace [\massArgnt{}]^{-1} \basisspace^T \stweightingMatArg{} +   \basisspace^T \stweightingMatArg{} \frac{d}{dt} \bigg] \bigg(  \basisspace \genstateDotArg{}{t}  -  \velocity(\veloargsrom) \bigg) = \bz.

@@

Pulling $\mass^{-1} \basisspace^T \stweightingMatArg{}$ inside the parenthesis on the second term,

@@

\basisspace^T \bigg[\frac{\partial \velocity}{\partial \state^n}(\veloargsrom) \bigg]^T \stweightingMatArg{}\bigg( \mathbf{I} - \basisspace[\massArgnt{}]^{-1} \basisspace^T \stweightingMatArg{} \bigg)  \bigg(  \basisspace \genstateDotArg{}{t}   -  \velocity(\veloargsrom) \bigg)  + \\  
\bigg[ \basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsrom) \bigg]^T \stweightingMatArg{} \basisspace  +   \mass \frac{d}{dt} \bigg] \bigg( \genstateDotArg{}{t}  - \mass^{-1} \basisspace^T \stweightingMatArg{}  \velocity(\veloargsrom) \bigg) = \bz.
